Query Optimization for Dynamic Imputation

### 2017-03-01 paper/conclusion.tex

\section{Conclusion}
Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective cost model to consider trade-offs in 
imputation quality and runtime. We show that different values for the trade-off parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with the \textit{CDC NHANES}, \textit{freeCodeCamp} and \textit{ACS} datasets
show that \ProjectName{} can be used successfully with real-world data, to run standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
 the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered (0-\highsmapealphaoneexacs\% error, \Cref{sec:results}). Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

We discuss the long history of dealing with missing data both in the statistical learning and database communities.
Similarly to existing work, we consider the impact of null values in databases and develop a simple set of invariants to 
successfully plan around them. In contrast to existing statistical learning work, our emphasis is not on the specific algorithm
used to impute but rather on the placement of imputation steps in the execution of a query. In contrast to existing database work,
we incorporate imputation into a cost-based optimizer and hide any details
regarding missing values inside the system, allowing users to use traditional SQL as if the database were complete.
While prior work has incorporated operations such as prediction into their databases, this has been domain-specific
in some cases, and in others, they have not integrated these operations into the planner. The novelty
of our contribution lies in the formalization of missing value imputation for query planning, which results in performance
gains over traditional approaches. This approach acknowledges that different users performing different queries on the same
dataset will likely have varying imputation needs and that execution plans should appropriately reflect that variation.

\subsection{Future Work}
\ProjectName{} opens up multiple avenues for further work. We sketch out some of these
directions.

\textbf{Imputation confidence:} Obtaining confidence measures for query results
produced when missing values are imputed with a machine learning model is
another possible improvement to the system. These measures could be obtained in general
using resampling methods like cross-validation and bootstrapping\cite{kohavi1995study}.
In contrast to standard confidence calculations, there is the added complexity that
imputation takes place at various points in the plan, so the composition of results
through standard query operations needs to be taken into account when computing confidence intervals.
Finally, multiple imputation could be considered explicitly and confidence measures like variances could
similarly propagate through the query plan.

>>>>

\section{Conclusion}
Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple yet effective cost model to consider trade-offs in 
imputation quality and runtime. We show that different values for the trade-off parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with the CDC NHANES, freeCodeCamp and ACS datasets
show that \ProjectName{} can be used successfully with real-world data to run standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
 the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered (0-\highsmapealphaoneexacs\% error, \Cref{sec:results}). Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

We discuss the long history of dealing with missing data both in the statistical learning and database communities.
Similar to existing work, we consider the impact of \nullv{} values in databases and develop a simple set of invariants to 
successfully plan around them. In contrast to existing statistical learning work, our emphasis is not on the specific algorithm
used to impute but rather on the placement of imputation steps in the execution of a query. In contrast to existing database work,
we incorporate imputation into a cost-based optimizer and hide any details
regarding missing values inside the system, allowing users to use traditional SQL as if the database were complete.
While prior work has incorporated operations such as prediction into their databases, this has been domain-specific
in some cases, and in others, they have not integrated these operations into the planner. The novelty
of our contribution lies in the formalization of missing value imputation for query planning, which results in performance
gains over traditional approaches. This approach acknowledges that different users performing different queries on the same
dataset will likely have varying imputation needs and that execution plans should appropriately reflect that variation.

\subsection{Future Work}
\ProjectName{} opens up multiple avenues for further work.

\textbf{Imputation confidence:} Obtaining confidence measures for query results
produced when missing values are imputed with a statistical model is
another possible improvement to the system. These measures could be obtained in general
using resampling methods like cross-validation and bootstrapping\cite{kohavi1995study}.
In contrast to standard confidence calculations, one added complexity is that
imputation takes place at various points in the plan, so the composition of results
through standard query operations needs to be taken into account when computing confidence
measures.
In addition, multiple imputation could be used within query execution and confidence measures like variances could
similarly propagate through the query plan.

### 2017-03-01 paper/introduction.tex

A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while mitigating the number of dropped rows. Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

>>>>

A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while reducing the number of dropped rows. Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.
